Predicting the effects of character displacement on food-web dynamics

### 2020-07-02 manuscript_ECD_model.Rmd
incorporate reviewer and editor comments.

where $a_{ij}$ is the attack rate of consumer $j$ on resource $i$. 
This first scenario is the starting point for many models of resource competition [@MacArthur1972]; however, it does not reflect many food webs where consumers are mobile and their foraging behavior links resources that occur in different habitats [@McCann2005]. 
The second scenario accounts for this spatial context (Fig. \ref{fig:foraging_scenarios}b) and takes the form:

@@

Given that character displacement occurred across these foraging scenarios, I focus here on its consequences for food-web dynamics.
To do this, I analyzed differences in resource abundances and food-web stability at equilibrium. 
An equilibrium is reached when there is no change in the population growth rates of consumers and resources (i.e., the rates of change in equation \ref{eq:1} are 0), and solving the system at this point gives equilibrium abundances for each resource ($\hat R_i$) and each consumer ($\hat C_j$). 
I also compared the local stability of these food webs using standard methods [@Otto2007]. 
This stability analysis derives the dominant eigenvalue, $\lambda_{\text{max}}$, of the matrix of partial derivatives of each species' population growth rate (given by equation \ref{eq:1}) with respect to each species' abundance evaluated at equilibrium. 
If $-\lambda_{max}>0$, then the food web will return to equilibrium after a small perturbation (i.e., it is locally stable), with more positive values indicating a faster return time.
If $-\lambda_{max}<0$, then the food web is not locally stable.

@@

To gain insight to the eco-evolutionary feedback generated by character displacement, I conducted simulations using an Adaptive Dynamics approach. 
Specifically, after letting consumer and resource abundances reach a steady state, I created a mutant consumer by randomly choosing one and modifying its attack rate on one resource by either subtracting or adding a small constant (0.01 in the following simulations) with equal probability. 
The mutant's attack rate on the other resource was determined by a tradeoff, such that $\big(a_{\text{1},j}/{A}\big)^n+\big(a_{\text{2},j}/{A}\big)^n=1$, where $A$ is the total investment in attack rates and $n$ describes the shape of the tradeoff [@Sargent2006]. 
This function has the useful property that it differentiates between cases where intermediate combinations of $a_{\text{1},j}$ and $a_{\text{2},j}$ are higher than the extremes (when $n>1$, green line in Fig. \ref{fig:tradeoff}) or, conversely, where the two extremes are higher than intermediate investments (when $n<1$, orange line in Fig. \ref{fig:tradeoff}). 
When $n=1$, the tradeoff function is linear, and all combinations of $a_{\text{1},j}$ and $a_{\text{2},j}$ have the same total attack rate (blue line in Fig. \ref{fig:tradeoff}). 
Assuming the mutant consumer was rare, I then determined whether the mutant had higher relative fitness than the resident consumer, and thus could invade and replace the resident consumer. 
If the mutant was able to invade, I updated the attack rate of the resident consumer to the mutant attack rate and allowed consumer and resource abundances to reach a steady state. 
I then repeated the simulation up to 10,000 times, which was sufficient for consumers to either reach an evolutionary stable strategy [ESS, @Smith1973] or an evolutionary limit (e.g., $\frac{a_{ij}}{a_{\text{1},j}+a_{\text{2},j}}$ is constrained to a maximum of 1 and minimum of 0). 
Unless otherwise noted, I conducted simulations with the following parameter values: *r* = 1; *K* = 4; *e* = 0.8; *m* = 1; *A* = 2; *h* = 0.4; and $w_{11} = w_{22} = 0.6$. 
I set an initial value of $a_{11} = a_{22} = 1.2$, while $a_{12}$ and $a_{21}$ depended on the value of *n*. 
I set initial consumer and resource abundances to: $R_1 = R_2 = 2$; $C_1 = C_2 = 1$.
All simulations were conducted in R [@R] and the code to reproduce these simulations is publically available on GitHub <!--(https://github.com/mabarbour/ECD_model) -->and has been archived with Zenodo <!-- (https://zenodo.org/badge/latestdoi/79349192)-->(please contact journal office for links).

@@

The effect of character displacement on resources corresponds to its impact on food-web stability. 
For example, when character displacement decreases resource abundances (orange points in Fig. \ref{fig:plot_fig3}b), there is also a decrease in food-web stability (Fig. \ref{fig:plot_fig3}c). 
Character displacement may not affect or even increase food-web stability (blue and green lines in Fig. \ref{fig:plot_fig3}c); however, evolution does not favor strong divergence in these scenarios (blue and green points in Fig. \ref{fig:plot_fig3}a), which dampens these contingent effects.
Note that the dip in stability occurs when both consumers evolve to be generalists, a situation that is not favored in any of the foraging scenarios I examined (Fig. \ref{fig:plot_fig3}c).

@@

As seen previously, the effect of character displacement on resource abundances qualitatively corresponds to its effect on food-web stability (Fig. \ref{fig:plot_fig3}f). 
Specifically, character displacement decreases food-web stability, regardless of the tradeoff in attack rates.
This is not simply a consequence of having an additional consumer in the system, but emerges from the eco-evolutionary feedback between character displacement and resource suppression (Fig. \ref{fig:plot_fig3}c). 
For example, when the tradeoff is concave up (orange), the initial two-consumer food web (small circle) is more stable than when there is only one consumer (small triangle); however, this pattern switches by the end of the eco-evolutionary simulation (large points).

@@

```{r plot_fig3, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:plot_fig3}**Effect of character displacement on food-web dynamics under different evolutionary tradeoffs and foraging scenarios.** Lines show predicted values when both consumers and resources are present. Different line colors correspond to different tradeoffs in attack rates (green, *n* = 1.15; blue, *n* = 1; orange, *n* = 0.85). Large circles (two consumers) and triangles (one consumer) correspond to the end points of the eco-evolutionary simulation for *C*~1~ (the choice to display *C*~1~ was arbitrary), whereas as small shapes correspond to the starting points (only in stability panels). In both foraging scenarios, feeding rates increase linearly with resource abundance, but the equation for the effective attack rate is different."}
knitr::include_graphics("Fig_3_MacArthur_LawlorSmith.pdf")
```

@@

In the first two foraging scenarios, character displacement influences food-web stability, but all of the food webs ultimately return to a stable equilibrium (because $-\lambda_{max}>0$, see Appendix S2-3). 
In this more realistic model, however, whether the food web is locally stable depends on consumer and resource parameters. 
Specifically, I found that the two-consumer food web will transition from having a locally stable equilibrium to a limit cycle under the following conditions (derived using Routh-Hurwitz criteria in Appendix S1):

@@

@@

@@

One of the criteria used to demonstrate ecological character displacement is that ``sites of sympatry [two consumers] and allopatry [one consumer] should not differ greatly in food $\textbf{[}$resource abundances$\textbf{]}$'' [@Schluter1992].
In contrast, my results indicate that character displacement causes predictable differences in resource abundances.
In fact, the ecological and evolutionary scenarios that favored the largest character displacement always decreased the relative abundance of resources. <!-- build paragraph around this topic sentence -->
For example, if mobile consumers compete for resources that occur in different habitats, then character displacement always resulted in lower resource abundances.<!--the magnitude of character displacement was highest when mobile consumers competed for resources that occur in different habitats.-->
Threespine stickleback, one of the classic examples of character displacement, exemplify this foraging scenario [@Schluter1992; @Schluter2000]. 
Stickleback must move between the pelagic and littoral zones of a lake when foraging for zooplankton and benthic invertebrates, respectively.
The theory developed here predicts that resource abundances will be lower in lakes where competing stickleback have undergone character displacement compared to lakes with only a single species of stickleback.
Interestingly, a disproportionate number of the documented cases of character displacement involve carnivores [@Schluter2000] that are larger, and likely more mobile, than their resources [@McCann2005], suggesting that many cases of ecological character displacement may result in lower resource availability.

@@

Similarly, the evolutionary tradeoff that favored character displacement decreased resource availability across all foraging scenarios.
Although data on the shape of the tradeoff in consumer foraging traits is scarce, two classic examples of character displacement, Darwin's finches and threespine stickleback, both appear to exhibit a tradeoff where extreme trait values increase the net foraging rate of consumers [@Schluter1985; @Arnegard2014]. 
While it is theoretically possible that character displacement does not alter (or even increase) resource abundances, this was limited to the simplest, and arguably least realistic, foraging scenario and under tradeoffs that did not favor large displacements, and thus less likely to detect in nature.
Taken together, my results call for empirical work to test these clear theoretical predictions and suggest a revision is needed for one of the criteria used to demonstrate character displacement.

<!-- consider removing entirely. This is a prediction I could test with Seth in another paper...
Character displacement is far from an all-or-nothing process, but may vary in its magnitude [@Schluter2000]. 
My results suggest that the magnitude of character displacement will be negatively correlated with total resource abundances, at least when resources occur in distinct habitats (line trajectories in \ref{fig:LS_Plot}B, \ref{fig:McCann_Plot}B).
This prediction could be tested either experimentally, by putting populations that vary in divergence in a common resource setting, or with field data that statistically controls for other factors affecting resources.
Note though, this prediction will only apply to comparisons within species where there is likely little difference in the shape of the tradeoff among populations.
-->

>>>>

where $a_{ij}$ is the attack rate of consumer $j$ on resource $i$. 
This first scenario is the starting point for many models of resource competition [@MacArthur1972]; however, it does not reflect many food webs where consumers are mobile and their foraging behavior links resources that occur in different habitats [@Holt1984]. 
The second scenario accounts for this spatial context (Fig. \ref{fig:foraging_scenarios}b) and takes the form:

@@

Given that character displacement occurred across these foraging scenarios, I focus here on its consequences for food-web dynamics.
To do this, I analyzed differences in resource abundances and food-web stability at equilibrium. 
An equilibrium is reached when there is no change in the population growth rates of consumers and resources (i.e., the rates of change in equation \ref{eq:1} are 0), and solving the system at this point gives equilibrium abundances for each resource ($\hat R_i$) and each consumer ($\hat C_j$). 
I also compared the local stability of these food webs using standard methods [@Otto2007]. 
This stability analysis derives the dominant eigenvalue, $\lambda_{\text{max}}$, of the matrix of partial derivatives of each species' population growth rate (given by equation \ref{eq:1}) with respect to each species' abundance evaluated at equilibrium. 
All eigenvalues must be negative for the food web to be locally stable; however, the dominant eigenvalue, $\lambda_{\text{max}}$, is the one closest to positive infinity and ultimately determines food-web stability. 
If $\lambda_{\text{max}}>0$, then the food web is not locally stable. 
If $\lambda_{\text{max}}<0$, then the food web will return to equilibrium after a small perturbation (i.e., it is locally stable), with more negative values indicating a faster return time (i.e., greater stability). 
I chose to plot $-\lambda_{\text{max}}$ rather than $\lambda_{\text{max}}$, so that increasing values would indicate greater stability, which I felt was a more intuitive way to present the results.
<!-- old test: If $-\lambda_{max}>0$, then the food web will return to equilibrium after a small perturbation (i.e., it is locally stable), with more positive values indicating a faster return time.
If $-\lambda_{max}<0$, then the food web is not locally stable.-->

@@

To gain insight to the eco-evolutionary feedback generated by character displacement, I conducted simulations using an Adaptive Dynamics approach. 
Specifically, after letting consumer and resource abundances reach a steady state, I created a mutant consumer by randomly choosing one and modifying its attack rate on one resource by either subtracting or adding a small constant (0.01 in the following simulations) with equal probability. 
The mutant's attack rate on the other resource was determined by a tradeoff, such that $\big(a_{\text{1},j}/{A}\big)^n+\big(a_{\text{2},j}/{A}\big)^n=1$, where $A$ is the total investment in attack rates and $n$ describes the shape of the tradeoff [@Sargent2006]. 
This function has the useful property that it differentiates between cases where intermediate combinations of $a_{\text{1},j}$ and $a_{\text{2},j}$ are higher than the extremes (when $n>1$, green line in Fig. \ref{fig:tradeoff}) or, conversely, where the two extremes are higher than intermediate investments (when $n<1$, orange line in Fig. \ref{fig:tradeoff}). 
When $n=1$, the tradeoff function is linear, and all combinations of $a_{\text{1},j}$ and $a_{\text{2},j}$ have the same total attack rate (blue line in Fig. \ref{fig:tradeoff}). 
Assuming the mutant consumer was rare, I then determined whether the mutant had higher relative fitness than the resident consumer, and thus could invade and replace the resident consumer. 
If the mutant was able to invade, I updated the attack rate of the resident consumer to the mutant attack rate and allowed consumer and resource abundances to reach a steady state. 
I then repeated the simulation up to 10,000 times, which was sufficient for consumers to either reach an evolutionary stable strategy [ESS, @Smith1973] or an evolutionary limit (e.g., $\frac{a_{ij}}{a_{\text{1},j}+a_{\text{2},j}}$ is constrained to a maximum of 1 and minimum of 0).

<!--As with any model, the Adaptive Dynamics approach makes certain assumptions. First, it assumes a separation of time scales between ecological and evolutionary dynamics, an assumption that is becoming less tenable [@Hairston2005; @Hendry2016]. Second, it assumes there is no intraspecific variation within interacting species, despite the fact that intraspecific variation is a common feature of natural populations and often has important ecological consequences [@Bolnick2011; @Des_Roches2018].-->

The Adaptive Dynamics approach I use here is one of several possible approaches for studying the eco-evolutionary dynamics of interacting species. A strength of this approach is that it enables me to gain analytical insight to the effects of character displacement in more realistic foraging scenarios.
This is much less tractable in quantitative genetic [@Taper1985; @McPeek2017] or explicit genetic [@Doebeli1996] models of character displacement, which is why the foraging scenarios previously examined have been limited [but see @McPeek2017]. 
A weakness, however, is that I assume a separation of time scales between ecological and evolutionary dynamics, an assumption that is becoming less tenable [@Hairston2005; @Hendry2016]. 
In addition, I assume that there is no intraspecific variation within interacting species, despite the fact that intraspecific variation is a common feature of natural populations and often has important ecological consequences [@Bolnick2011; @Des_Roches2018].
Still, models that explicitly include resource dynamics inevitably show that resource competition results in character displacement, regardless of whether a quantitative genetic or Adaptive Dynamics approach is used [@Lawlor1976; @Taper1985].
Therefore, it should still serve as a good starting point for studying the ecological consequences of character displacement (but see \textbf{Caveats} section in the discussion).

Unless otherwise noted, I conducted simulations with the following parameter values: *r* = 1; *K* = 4; *e* = 0.8; *m* = 1; *A* = 2; *h* = 0.4; and $w_{11} = w_{22} = 0.6$. 
I set an initial value of $a_{11} = a_{22} = 1.2$, while $a_{12}$ and $a_{21}$ depended on the value of *n*. 
I set initial consumer and resource abundances to: $R_1 = R_2 = 2$; $C_1 = C_2 = 1$.
All simulations were conducted in R [@R] and the code to reproduce these simulations is publically available on GitHub <!--(https://github.com/mabarbour/ECD_model) -->and has been archived with Zenodo <!-- (https://zenodo.org/badge/latestdoi/79349192)-->(please contact journal office for links).

@@

The effect of character displacement on resources corresponds to its impact on food-web stability. 
For example, when character displacement decreases resource abundances (orange points in Fig. \ref{fig:plot_fig3}b), there is also a decrease in food-web stability (Fig. \ref{fig:plot_fig3}c). 
Character displacement may not affect or even increase food-web stability (blue and green lines in Fig. \ref{fig:plot_fig3}c); however, evolution does not favor strong divergence in these scenarios (blue and green points in Fig. \ref{fig:plot_fig3}a), which dampens these contingent effects.
Note that the dip in stability toward a dominant eigenvalue of zero is an artifact of the two consumers becoming ecologically identical ($a_{11}=a_{21}=a_{12}=a_{22}$) on identical resources, and effectively become neutral species [@McPeek2019Neutral].
Although neutrality is an ecological possibility, it is never favored by evolution in these models since character displacement occurred across all the foraging scenarios I examined.<!-- never occurs this situation is not favored in any of the foraging scenarios I examined.
Note that the dip in stability occurs when both consumers evolve to be generalists, a situation that is not favored in any of the foraging scenarios I examined (Fig. \ref{fig:plot_fig3}c).
In fact, if the consumers were ecologically identical ($a_{1,j}=a_{2,j}$), then they are neutral species, as indicated by the dominant eigenvalue becoming zero [@McPeek2019Neutral]. 
This only occurs -->

@@

As seen previously, the effect of character displacement on resource abundances qualitatively corresponds to its effect on food-web stability (Fig. \ref{fig:plot_fig3}f). 
Specifically, character displacement decreases food-web stability, regardless of the tradeoff in attack rates.
This is not simply a consequence of having an additional consumer in the system, but emerges from the eco-evolutionary feedback between character displacement and resource suppression (Fig. \ref{fig:plot_fig3}c). 
For example, when the tradeoff is concave up (orange), the initial two-consumer food web (small circle) is more stable than when there is only one consumer (small triangle); however, this pattern switches by the end of the eco-evolutionary simulation (large points).
As with the previous model, the dip in stability corresponds to the situation where consumers are effectively identical ($w_{11}a_{11}=w_{21}a_{21}=w_{12}a_{12}=w_{22}a_{22}$) and become neutral species, which is never favored in these models.
<!--Note that consumer neutrality only becomes possible with certain specific combinations of parameters in these models (e.g., $a_{1,j}=a_{2,j}$ or $w_{1,j}a_{1,j}=w_{2,j}a_{2,j}$).-->

It is worth noting that the scales are different between the left and right columns of Figure \ref{fig:plot_fig3}. When resources occur in different habitats (relative to the same habitat), the consumer's effective attack rate is lower because it is multiplied by the proportion $w_{ij}$. This results in relatively higher resource abundances and more stable food webs compared to when resources occur in the same habitat, and corresponds to the inherent stabilizing role of spatial heterogeneity in food webs [@Holt1984; @McCann2005].

@@

```{r plot_fig3, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:plot_fig3}**Effect of character displacement on food-web dynamics under different evolutionary tradeoffs and foraging scenarios.** Lines show predicted values when both consumers and resources are present. Different line colors correspond to different tradeoffs in attack rates (green, *n* = 1.15; blue, *n* = 1; orange, *n* = 0.85). Large circles (two consumers) and triangles (one consumer) correspond to the end points of the eco-evolutionary simulation for *C*~1~ (the choice to display *C*~1~ was arbitrary), whereas small shapes correspond to the starting points (only in stability panels). In both foraging scenarios, feeding rates increase linearly with resource abundance, but the equation for the effective attack rate is different."}
knitr::include_graphics("Fig_3_MacArthur_LawlorSmith.pdf")
```

```{r normalize-fig-3, include=FALSE}
## setup normalization
# from https://www.r-bloggers.com/custom-axis-transformations-in-ggplot2/
require(scales) # trans_new() is in the scales library
norm_Mac <- bind_rows(MacArthur_2C_2R_tradeoff_dynamics, MacArthur_1C_2R_tradeoff_dynamics) %>%
  mutate(eff_a = a11 + a12,
         RT = R1 + R2,
         Stability = -1*max.Re.eigen)
# easiest way to manually do this in the aesthetics

norm_LS <- bind_rows(LawlorSmith_2C_2R_tradeoff_dynamics, LawlorSmith_1C_2R_tradeoff_dynamics) %>%
  mutate(eff_a = w11*a11 + w12*a12,
         RT = R1 + R2,
         Stability = -1*max.Re.eigen)

## normalize plots MacArthur
Mac_attackNorm <- ggplot(tradeoff_df, aes(x=aii/(aii+aij), y=(aii+aij - min(norm_Mac$eff_a))/(max(norm_Mac$eff_a) - min(norm_Mac$eff_a)), color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(MacArthur_displacements, Time=="End"), aes(x = a11/(a11+a12), y=(a11+a12 - min(norm_Mac$eff_a))/(max(norm_Mac$eff_a) - min(norm_Mac$eff_a)), fill=factor(n), shape=Competitor), size=3, color="white") + 
  ylab(expression(a[ii]+a[ij])) +
  scale_shape_manual(values=c(24,21)) +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  guides(fill = guide_legend(override.aes=list(shape=21)))

Mac_resourcesNorm <- cbind.data.frame(select(tradeoff_df, n, aii, aij), MacArthur_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=(R1+R2 - min(norm_Mac$RT))/(max(norm_Mac$RT) - min(norm_Mac$RT)), color = factor(n))) + 
  geom_line() +
  geom_line(data = cbind.data.frame(select(tradeoff_df, n, aii, aij), MacArthur_1C_2R_tradeoff_dynamics), alpha = 0.1) +
  geom_point(data = filter(MacArthur_displacements, Time=="End"), aes(x = a11/(a11+a12), y=(R1+R2 - min(norm_Mac$RT))/(max(norm_Mac$RT) - min(norm_Mac$RT)), fill=factor(n), shape=Competitor), size=3, color="white") +
  ylab(expression(R[i]+R[j])) +
  scale_shape_manual(values=c(24,21)) +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  guides(fill = guide_legend(override.aes=list(shape=21)))

Stability_MacNorm <- cbind.data.frame(select(tradeoff_df, n, aii, aij), MacArthur_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=(-1*max.Re.eigen - min(norm_Mac$Stability))/(max(norm_Mac$Stability) - min(norm_Mac$Stability)), color = factor(n))) + 
  geom_line() +
  geom_line(data = cbind.data.frame(select(tradeoff_df, n, aii, aij), MacArthur_1C_2R_tradeoff_dynamics), alpha = 0.1) +
  geom_point(data = filter(MacArthur_displacements, Time=="End"), aes(x = a11/(a11+a12), y=(-1*max.Re.eigen - min(norm_Mac$Stability))/(max(norm_Mac$Stability) - min(norm_Mac$Stability)), fill=factor(n), shape=Competitor), size=3, color="white", show.legend=FALSE) + 
  geom_point(data = filter(MacArthur_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=(-1*max.Re.eigen - min(norm_Mac$Stability))/(max(norm_Mac$Stability) - min(norm_Mac$Stability)), fill=factor(n), shape=Competitor), size=1.5, color="white") +
  ylab(expression("Stability "(-lambda[max]))) + 
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  guides(fill = guide_legend(override.aes=list(shape=21)))

MacsNorm <- plot_grid(
  Mac_attackNorm + xlab("") + ylab(bquote(atop("Effective attack rate","("*a[1*","*j]+a[2*","*j]*")"))) +
    expand_limits(y = c(0,1)) +
    theme_cowplot() +
    theme(legend.position = "none", axis.text = element_text(size=10), axis.text.x = element_blank(), plot.title = element_text(size = 10), axis.title.y=element_text(vjust=0, size=12)), 
  Mac_resourcesNorm + xlab("") + ylab(bquote(atop("Resource abundance","("*R[1]+R[2]*")"))) + 
    expand_limits(y = c(0,1)) +
    theme_cowplot() +
    theme(legend.position = "none", axis.text = element_text(size=10), axis.text.x = element_blank(), axis.title.y=element_text(vjust=0, size=12)), 
  Stability_MacNorm + scale_x_continuous(labels = c(0,0.25,0.5,0.75,1)) + xlab("") + ylab(bquote(atop("Food-web stability","("*-lambda[max]*")"))) + 
    expand_limits(y = c(0,1)) +
    theme_cowplot() +
    theme(legend.position = "none", axis.text = element_text(size=10), axis.title.y=element_text(vjust=0, size=12)), 
  ncol=1, align='v', labels = c("","b)","c)"), hjust = -4, vjust = 0.5, label_size = 10
)
#Macs_title <- ggdraw() + draw_label("Resources: Same Habitat", fontface = 'bold', size = 10, hjust = 0.2)
plot_MacsNorm <- plot_grid(Macs_title, MacsNorm, ncol = 1, rel_heights = c(0.1, 1))

## LawlorSmith
LS_attackNorm <- ggplot(tradeoff_df, aes(x=aii/(aii+aij), y=(wii*aii+wij*aij - min(norm_LS$eff_a))/(max(norm_LS$eff_a) - min(norm_LS$eff_a)), color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(LawlorSmith_displacements, Time=="End"), aes(x = a11/(a11+a12), y=(w11*a11+w12*a12 - min(norm_LS$eff_a))/(max(norm_LS$eff_a) - min(norm_LS$eff_a)), fill=factor(n), shape=Competitor), size=3, color="white") + 
  #geom_point(data = filter(LawlorSmith_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=w11*a11+w12*a12, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  ylab(expression(w*a[ii]+(1-w)*a[ij])) +
  #xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_shape_manual(values=c(24,21)) +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  guides(fill = guide_legend(override.aes=list(shape=21)))

LS_resourcesNorm <- cbind.data.frame(select(tradeoff_df, n, aii, aij), LawlorSmith_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=(R1+R2 - min(norm_LS$RT))/(max(norm_LS$RT) - min(norm_LS$RT)), color=factor(n))) + 
  geom_line() +
  geom_line(data = cbind.data.frame(select(tradeoff_df, n, aii, aij), LawlorSmith_1C_2R_tradeoff_dynamics), alpha = 0.1) +
  geom_point(data = filter(LawlorSmith_displacements, Time=="End"), aes(x = a11/(a11+a12), y=(R1+R2 - min(norm_LS$RT))/(max(norm_LS$RT) - min(norm_LS$RT)), fill=factor(n), shape=Competitor), size=3, color="white") + 
  #geom_point(data = filter(LawlorSmith_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=R1+R2, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  ylab(expression(R[i]+R[j])) +
  #xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_shape_manual(values=c(24,21)) +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  guides(fill = guide_legend(override.aes=list(shape=21)))

Stability_LSNorm <- cbind.data.frame(select(tradeoff_df, n, aii, aij), LawlorSmith_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=(-1*max.Re.eigen - min(norm_LS$Stability))/(max(norm_LS$Stability) - min(norm_LS$Stability)), color = factor(n))) + 
  geom_line() +
  geom_line(data = cbind.data.frame(select(tradeoff_df, n, aii, aij), LawlorSmith_1C_2R_tradeoff_dynamics), alpha = 0.1) +
  geom_point(data = filter(LawlorSmith_displacements, Time=="End"), aes(x = a11/(a11+a12), y=(-1*max.Re.eigen - min(norm_LS$Stability))/(max(norm_LS$Stability) - min(norm_LS$Stability)), fill=factor(n), shape=Competitor), size=3, color="white", show.legend=FALSE) +
  geom_point(data = filter(LawlorSmith_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=(-1*max.Re.eigen - min(norm_LS$Stability))/(max(norm_LS$Stability) - min(norm_LS$Stability)), fill=factor(n), shape=Competitor), size=1.5, color="white") +
  ylab(expression("Stability "(-lambda[max]))) + 
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  guides(fill = guide_legend(override.aes=list(shape=21)))

LSsNorm <- plot_grid(
  LS_attackNorm  + xlab("") + ylab(bquote(w[1*","*j]*a[1*","*j]+w[2*","*j]*a[2*","*j])) + 
    expand_limits(y = c(0,1)) +
    theme_cowplot() +
    theme(legend.position = "none", axis.text = element_text(size=10), axis.text.x = element_blank(), plot.title = element_text(size = 10), axis.title.y=element_text(size=12)), 
  LS_resourcesNorm + xlab("") + ylab("") +
    expand_limits(y = c(0,1)) +
    theme_cowplot() +
    theme(legend.position = "none", axis.text = element_text(size=10), axis.text.x = element_blank()),   
  Stability_LSNorm + scale_x_continuous(labels = c(0,0.25,0.5,0.75,1)) + xlab("") + ylab("") + 
    expand_limits(y = c(0,1)) +
    theme_cowplot() +
    theme(legend.position = "none", axis.text = element_text(size=10)), 
  ncol=1, align='v', labels = c("","e)","f)"), hjust = -1.5, vjust = 0.5, label_size = 10
)
#LSs_title <- ggdraw() + draw_label("Resources: Different Habitat", fontface = 'bold', size = 10, hjust = 0.3)
plot_LSsNorm <- plot_grid(LSs_title, LSsNorm, ncol = 1, rel_heights = c(0.1, 1))

plot_fig3Norm <- plot_grid(plot_MacsNorm, plot_LSsNorm, ncol=2, rel_widths = c(1,0.9)) 
plot_fig3_labelNorm <- plot_fig3Norm + 
  draw_label("a)", x = 0.098, y = 0.925, fontface = 'bold', size = 10) + 
  draw_label("d)", x = 0.566, y = 0.925, fontface = 'bold', size = 10)
plot_fig3_subNorm <- add_sub(plot_fig3_labelNorm, expression(paste("Specialization ", bgroup("(",frac(a[ij],a[1*","*j]+a[2*","* j]),")"))), vjust = 0, size=12)

save_plot(filename="Fig_3_MacArthur_LawlorSmith_Normalized.pdf", plot=plot_fig3_subNorm, base_height = 6.5, base_width = 6)
```

@@

In the first two foraging scenarios, character displacement influences food-web stability, but all of the food webs ultimately return to a stable equilibrium (because $-\lambda_{max}>0$, see Appendix S2-3). 
In this more realistic model, however, whether the food web is locally stable depends on consumer and resource parameters. 
Specifically, I found that the two-consumer food web will become unstable<!--transition from having a locally stable equilibrium to a limit cycle--> under the following conditions (derived using Routh-Hurwitz criteria in Appendix S1):

@@

```{r check-negative-eigenvalues, include=FALSE}
# MacArthur
filter(MacArthur_2C_2R_tradeoff_dynamics, a11/(a11+a12) == 0.5) %>%
  select(a11,a12,max.Re.eigen) %>%
  knitr::kable()

filter(MacArthur_2C_2R_tradeoff_dynamics, max.Re.eigen == 0) %>%
  select(a11,a12,max.Re.eigen) %>%
  knitr::kable()

with(MacArthur_2C_2R_tradeoff_dynamics, table(max.Im.eigen > 0)) %>% knitr::kable()

with(MacArthur_1C_2R_tradeoff_dynamics, table(max.Im.eigen > 0)) %>% knitr::kable()

# LawlorSmith
filter(LawlorSmith_2C_2R_tradeoff_dynamics, w11*a11/(w11*a11+w12*a12) == 0.5) %>%
  select(a11,a12,w11,w12,max.Re.eigen) %>%
  knitr::kable()

filter(LawlorSmith_2C_2R_tradeoff_dynamics, max.Re.eigen == 0) %>%
  select(a11,a12,w11,w12,max.Re.eigen) %>%
  knitr::kable()

with(LawlorSmith_2C_2R_tradeoff_dynamics, table(max.Im.eigen > 0)) %>% knitr::kable()

with(LawlorSmith_1C_2R_tradeoff_dynamics, table(max.Im.eigen > 0)) %>% knitr::kable()

with(McCann_2C_2R_tradeoff_dynamics, table(max.Im.eigen > 0)) %>% knitr::kable()

with(McCann_1C_2R_tradeoff_dynamics, table(max.Im.eigen > 0)) %>% knitr::kable()
```

@@

<!-- stickleback paper about no differences in habitat: @Ormond2011. Interestingly, the species pair appears to be associated with lakes with low fish diversity...not clear with the trophic linkages are there -->

@@

One of the criteria used to demonstrate ecological character displacement is that ``sites of sympatry [two consumers] and allopatry [one consumer] should not differ greatly in food $\textbf{[}$resource abundances$\textbf{]}$, climate, or other environmental features affecting the phenotype'' [@Schluter1992]. <!-- added full criteria -->
In contrast, my results indicate that character displacement causes predictable differences in resource abundances.
In fact, the ecological and evolutionary scenarios that favored the largest character displacement always decreased the relative abundance of resources. <!-- build paragraph around this topic sentence -->
For example, if mobile consumers compete for resources that occur in different habitats, then character displacement always resulted in lower resource abundances.<!--the magnitude of character displacement was highest when mobile consumers competed for resources that occur in different habitats.-->
Threespine stickleback, one of the classic examples of character displacement, exemplify this foraging scenario [@Schluter1992; @Schluter2000]. 
Stickleback must move between the pelagic and littoral zones of a lake when foraging for zooplankton and benthic invertebrates, respectively.
The theory developed here predicts that resource abundances will be lower in lakes where competing stickleback have undergone character displacement compared to lakes with only a single species of stickleback.
Interestingly, a disproportionate number of the documented cases of character displacement involve carnivores [@Schluter2000] that are larger, and likely more mobile, than their resources [@McCann2005], suggesting that many cases of ecological character displacement may result in lower resource availability.

@@

Similarly, the evolutionary tradeoff that favored character displacement decreased resource availability across all foraging scenarios.
Although data on the shape of the tradeoff in consumer foraging traits is scarce, two classic examples of character displacement, Darwin's finches and threespine stickleback, both appear to exhibit a tradeoff where extreme trait values increase the net foraging rate of consumers [@Schluter1985; @Arnegard2014]. 
While it is theoretically possible that character displacement does not alter (or even increase) resource abundances, this was limited to the simplest, and arguably least realistic, foraging scenario and under tradeoffs that did not favor large displacements, and thus less likely to detect in nature.
<!--Taken together, my results call for empirical work to test these clear theoretical predictions and suggest a revision is needed for one of the criteria used to demonstrate character displacement.-->

<!--Testing the prediction that character displacement decreases resource abundances could be done either experimentally, by putting consumers that have or have not undergone character displacement in a common resource setting, or with field data that statistically controls for other factors affecting resources.

Testing this prediction 
The most straightforward prediction to test is that character displacement decreases resource abundances.-->
